---
title: "Unix Tutorials"
author: "Abu Nayeem"
date: "December 24, 2014"
output: html_document
---

##### Introduction of UNIX

![To learn more](http://freeengineer.org/learnUNIXin10minutes.html)

UNIX is an operating system and has a graphical user interface (GUI). UniX is made up of three parts: *the kernal, the shell, and the programs*

**The Kernal**
This is the main hub of hte operating system: it handles time and memory to prorams and handle the filestor and and communications. It essentially acts like a server

**The Shell**
The shell acts as an interface between the user and kernal.The shell is a command line interpretator (CLI), and it essentially treats them as programs; the shell store all the previous commands.

**The Programs**
Everything in UNIX is either a process or file or process. It has a system heirarchy where (/) is the root diagram

##### Simple commands

`ls -a`: it lists all the files in the hidden directory
`cd ~`: change back to home directory

1) Copying files into current directory 
`cp filepath namefile`; Note that the directory needs to be writable
`mv file1 file2`; this moves the file, note you cna remname it
`rm file`; also rmdir removes a directory

2) Viewing text
`cat file.txt`; displays entire text
`less file.txt`; displays text up to a page
`head file.txt`; displays first ten lines
`tail file.txt`; displays last ten lines

3) Search contents of the file
`grep -i science file.txt`; this searches the word science and print the line regardless of capitalization
`grep -i 'science help' file.txt`; this is a pattern  
*Options:* `-v` (display do not match); `-n` (denote line number); `-c` print out only the total count of matched lines
*Multiple options:* `grep -ivc science file.txt`

4) WordCount
`wc -w science.txt`: count the total number of times science was used

5) Make a list; *press Control D to end the list/file*
```{r}
cat > list1
pear 
banana
apple
# Display the list created
cat list1 
```

Appending a list: close as normal
```{r}
cat >> list1
peach 
orange
```

Combine two documents:
`cat list1 list2 > biglist`; combines the two lists together into one

6) Manipulating lists [NOTE: they use a different symbol]

The `<` symbol redirects the input. For example:
`sort < biglist` ; sorts the elements in the list
`sort < biglist > sortlist` ; the following diagram sorts out biglist and save it under a different file

7) Find and list the number of users; *We use pipes, which are vertical lines*
`who | sort` ; list name of users and sort
`who | wc -1` ; find how many users are logged on

**Exercise**: Using pipes, diaplay all line sof list1 and list2 containing the letter 'p', and sort the result:
`cat list1 list2 | grep p | sort`

8) **Wildcards** use the asterick mark and designate any match

`ls list*`; this list any files that start with list
`ls *list`; list all the files that end with list
`ls ?list`; this matches *exactly one character*;  so yea to blist and not to brlist 

##### File System Security

`drwxrw-r-- 1 ee51ab 2450 Sept29 11:52 file1`: So the user is ee51ab, size is 2450, time created, and file name. The first section is the permission code, where `d` stands for directory

**Access rights of Files Terminology:**
`r (or -)`- indicates read permission where you can *read it and copy it*
`w`- indicates write permission, that is *permission to change a file.*
`x`- this allows the *permission to execute the file*

**Access rights of Directory Terminology:**
- `r` allows users to list files in the directory
- `w` users may delete files fromt he directory or move files into it
- `x` means the right to access the directory, granted you have permission

The groups of three:
- the *left group* gives the file permissions of the person who owns the file
- the *middle group* gives person in group the permission of the file
- the *right group* gives the permission from all others

**Examples:**
`-rwxrwxrwx` : a file that everyone can read, write, and execute
`-rw-------` : a file that only the owner can read and write, and no one else

**Changing Access Rights using chmod [Codes]**
- `u`: user ; `g`: group ; `o`:other ; `a`: all
- `r`: read ; `w`: write (and delete) ; `x`: execute (and access directory)
- `+`: add permission ; `-`: remove permission

*Examples:*
`chmod go-rwx biglist`: this states group/other lose permission read/write/execute
`chmod a+rw biglist`: this gives all users permission to read and write

**Note:** You can set up background processes, or hidden processes

##### Other UNIX commands

`df`: determines how much space left on the file system
`gzip science.txt`: compress the files to reduce the space
`gunzip science.txt.gz`: expand the zip files
`diff`: finds the difference of files
`find`: find files from the current directory(.)
`find . -name "*.txt" -print`: print all the text, and print the files

##### Compiling UNIX Software Programs

*Steps to install the software:*
- Locate and download the source code
- Unpack the source code
- Compile the code
- Install the resulting executable
- Set paths to the installation directory

*Simplest way to compile a package:*
- `cd` to the directory containing the packages's source code
- `./configure` to configure the package for your system
- Type `make` to compile the language
- Optionally, type `make check` to run any self-tests that come with the package
- Type `make stall` to install the programs and any data files and documentation
- Optionally, type `make clean` to remove the program binaries and object files from source code directory

THERE ARE PLENTY of MORE TOOLS in UNIX packages

##### UNIX Variables

There are two categories of variables. There is *shell variables* which are for temporary usage [undercased] and *environmental variables* are more greater permanence; they are usually capitalized. 

Environmental Variables Ex:
`echo $OSTYPE` ; display the current OS TYPE
To see all the values of Environmental variables place this command
`printenv | less`
To see all the shell variables 
`set | less`

When logging in to a UNIX host, two files are used **.login**, which stores the environmental variables and **.cshrc**, which stores the shell variables of current session.  **WARNING**: Never put command that run graphical displays on these files.

*Setting shell variables*
1) We can set the shell history to 200 by command: `set history = 200`. This command is changed temporarily
2) To **permanently** set value of history:
```{r}
nedit ~/.cshrc # a special editor to edit files
set history = 200
source .cshrc # this save the file
```

**Setting the Path!**

Notice: we often see `Command not found`. This can can happen by two means: 1) the command doesn;t exist or 2) we are not in *right path*

To correct the second issue, we need either manually set the path:
`set path = ($path ~/units174/bin)`
OR more simply *change the directory where the path works*



















`



